---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastRG

<!-- badges: start -->
[![R-CMD-check](https://github.com/RoheLab/fastRG/workflows/R-CMD-check/badge.svg)](https://github.com/RoheLab/fastRG/actions)
[![Codecov test coverage](https://codecov.io/gh/RoheLab/fastRG/branch/main/graph/badge.svg)](https://codecov.io/gh/RoheLab/fastRG?branch=main)
<!-- badges: end -->

`fastRG` quickly samples a broad class of network models known as generalized random dot product graphs (GRDPGs). In particular, for matrices $X$, $S$ and $Y$, `fastRG` samples a matrix $A$ with expectation $X S Y^T$ where the entries are independently Poisson distributed conditional on $X$ and $Y$. This is primarily useful when $A$ is the adjacency matrix of a graph. Crucially, the sampling is $\mathcal O(m)$, where $m$ is the number of the edges in graph, as opposed to the naive sampling approach, which is $\mathcal O(n^2)$, where $n$ is the number of nodes in the network. For additional details, see the [paper](https://arxiv.org/abs/1703.02998).

`fastRG` has two primary use cases:

  1. Sampling enormous sparse graphs that cannot feasibly be sampled with existing samplers, and
  1. validating new methods for random dot product graphs (and variants).
    
`fastRG` makes the latent parameters of random dot product graphs readily available to users, such that simulation studies for community detection, subspace recovery, etc, are straightforward.

## Installation

`fastRG` is not yet on CRAN. You can install the development version
with:

``` r
# install.package("devtools")
devtools::install_github("RoheLab/fastRG")
```

## Usage

There are two stages to sampling from generalized random dot product graphs. First, we sample the latent factors $X$ and $Y$. Then we sample $A$ conditional on those latent factors. `fastRG` mimics this two-stage sample structure. For example, to sample from a stochastic blockmodel, we first create the latent factors. 

```{r}
library(fastRG)

set.seed(27)

sbm <- sbm(n = 1000, k = 5, expected_density = 0.01)
```

You can specify the latent factors and the mixing matrix $B$ yourself, but there are also defaults to enable fast prototyping. Here $B$ was randomly generated with `Uniform[0, 1]` entries and nodes were assigned randomly to communities with equal probability of falling in all communities. Printing the result object gives us some additional information:

```{r}
sbm
```
Now, conditional on this latent representation, we can sample graphs. `fastRG` supports several different output types, each of which is specified by the suffix to `sample_*()` functions. For example, we can obtain an edgelist in a `tibble` with:

```{r}
sample_edgelist(sbm)
```

but we can just as easily obtain the graph as a sparse matrix

```{r}
A <- sample_sparse(sbm)
A[1:10, 1:10]
```

or an igraph object

```{r}
sample_igraph(sbm)
```

Note that every time we call `sample_*()` we draw a new sample.

```{r}
A <- sample_sparse(sbm)
B <- sample_sparse(sbm)

all(A == B)  # random realizations from the SBM don't match!
```

## Efficient spectral decompositions

If you would like to obtain the singular value decomposition of the population adjacency matrix conditional on latent factors, that is straightforward:

```{r}
s <- eigs_sym(sbm)
s$values
```

Note that eigendecompositions and SVDS (for directed graphs) use `RSpectra` and do not require explicitly forming large dense population adjacency matrices; the population decompositions should be efficient in both time and space for even large graphs.

## Key sampling options

There are several essential tools to modify graph sampling that you should know about. First there are options that affect the latent factor sampling:

- `expected_degree`: Set the expected average degree of the graph by scaling sampling probabilities. We *strongly, strongly* recommend that you always set this option. If you do not, it is easy accidentally sample from large and dense graphs.

- `expected_density`: Set the expected density of the graph by scaling sampling probabilities. You cannot specify both `expected_degree` and `expected_density` at the same time.

In the second stage of graph sampling, the options are:

- `poisson_edges`: Either `TRUE` or `FALSE` depending on whether you would like a Bernoulli graph or a Poisson multi-graph. Scaling via `expected_degree` assumes a Poisson multi-graph, with some limited exceptions.

- `allow_self_edges`: Whether nodes should be allowed to connect to themselves. Either `TRUE` or `FALSE`.

## Known issues

Sampling blockmodels with very small numbers of nodes (or blockmodels with the number of blocks `k` on the same order as `n`) results in a degeneracy that can cause issues.

## Related work

[`igraph`](https://igraph.org/r/) allows users to sample SBMs (in $\mathcal O(m + n + k^2)$ time) and random dot product graphs (in $\mathcal O(n^2 k)$ time). You can find the original research code associated with `fastRG` [here](https://github.com/raningtky/sampleRDPG). There is also a Python translation of the original code in Python [here](https://github.com/yunjhongwu/matrix-routines/blob/master/fastRG.py). Both of these implementations are bare bones.

