---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastRG

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

`fastRG` quickly samples a broad class of network models known as generalized random product graphs. In particular, for matrices `X`, `S` and `Y`, `fastRG` samples a matrix `A` with expectation `X S Y^T` where individual entries are Poisson distributed. We recommend that you think of `A` as the adjacency matrix for a graph (or a multi-graph). Crucially, the sampling is `O(m)`, where `m` is the number of the edges in graph. Other algorithms are `O(n^2)`, where `n` is the number of nodes in the network. For additional details, please have a look at the [paper](https://arxiv.org/abs/1703.02998).

## Installation

`fastRG` is not yet on CRAN. You can install the development version
with:

``` r
# install.package("devtools")
devtools::install_github("RoheLab/fastRG")
```

There's also a [Python implementation](https://github.com/yunjhongwu/matrix-routines/blob/master/fastRG.py).

## Example Usage

The easiest way to use `fastRG` is to use wrapper functions that sample from population graph models. For example, to sample from an Erdos-Renyi graph `n = 1,000,000` nodes and expected degree 5, we can use the `er()` function.

```{r}
library(fastRG)

A <- erdos_renyi(n = 10^6, avgDeg = 5)
```

By default we always get a `Matrix::sparseMatrix()`, but we can also ask for the graph as an edgelist as well. This results in a fast way to create `igraph` objects using `igraph::graph_from_edgelist()`.

```{r}
el <- erdos_renyi(n = 1000, avgDeg = 5, returnEdgeList = TRUE)
head(el)
```

```{r}
g <- igraph::graph_from_edgelist(el)
g
```
